\MakeUppercase{Layered Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-22 k-planar.tex
Pffft

@@

We make use of the following lemma due to \cite{pilipczuk.sieberts:polynomial-arxiv,pilipczuk.sieberts:polynomial-soda}
\begin{lem}\lemlabel{p-centered-treewidth}
  Any graph $H$ of treewidth at most $t$ has $\chi_p(H)\in O(p^t)$.
\end{lem}

@@

    \item $\gamma(v)=\gamma(x)$ where $\gamma:V(H)\to\{1,\ldots,O(p^t)\}$ is a $p$-centered colouring of $H$.
  \end{enumerate}
  Condition 1 is easy to satisfy since $|V_i\cap B_x|\le \ell$.  Condition~2 is straightforward.  Condition~3 is possible to satisfy thanks to \lemref{p-centered-treewidth}.  To see why this is a $p$-centered colouring, consider some $X\subseteq V(G)$ such that $G[X]$ is connected.
  \begin{enumerate}
    \item If there exists $v,w\in X$ with $v\in V_i$ and $w\in V_j$ with $j-i\ge p$ then, since $G[X]$ is connected, $G[X]$ contains a path from $v$ to $w$.  This path contains at least one vertex from $V_{i'}$ for each $i'\in\{i,i+1,\ldots,j\}$. Therefore, $|\{\beta(v'):v'\in X\}|\ge j-i+1 > p$, so $X$ receives more than $p$ distinct colours.

    \item Otherwise, $X\subseteq V_{i},\ldots,V_{i+s}$ for some $s<p$.  Let $H'=H[\{x\in V(H):B_x\cap X\neq\emptyset]$.  If $|\{\gamma(x):x\in V(H')\}| > p$ then $|\{\gamma(v):v\in X\}|> p$ so $|\{\phi(v):v\in X\}|> p$ and we are done.  Otherwise, since $\gamma$ is a $p$-centered colouring of $H$, there must exist some $x\in V(H')$ such that $\gamma(x)\neq\gamma(y)$ for every $y\in V(H')\setminus\{x\}$.
    For any $v,w\in B_x$ with $v\neq w$, either $v,w\in V_{i'}$ for some $i'\in\{i,i+1,\ldots,i+s\}$ in which case $\alpha(v)\neq\alpha(w)$; or $v\in V_{i'}$ and $w\in V_{i''}$ with $|i'-i''|< p$, in which case $\beta(v)\neq\beta(w)$. Therefore every vertex $v\in B_x$ receives a colour $\phi(v)$ distinct from every colour in $\{\phi(z):z\in X\setminus\{x\}\}$. 
  \end{enumerate}
\end{proof}

@@

Some possibilities
\begin{enumerate}
  \item $p$-centered colourings;
  \item clustered colouring;
  \item non-repetitive chromatic number;
  \item connection with queue layouts;
  \item $d$-map graphs, which are $d^2$-planar;
  \item connection with low-treewidth/treedepth colourings;
  \item An important open question is whether $k$-planar graphs have layered $H$-partitions whose layered width is upper bounded by a function of $k$ but in which the treewidth of $H$ is bounded by an absolute constant, independent of $k$.
  \item Are our current $H$-partitions $K_{3,s}$-free for some $s=s(k)$?  If so, this is good news for clustered colouring.
\end{enumerate}

>>>>

Prior to the current work, the strongest upper bound on the non-repetitive chromatic number of $n$-vertex planar $k$-planar graphs was $O(k\log n)$ \cite{dujmovic.morin.ea:layered}.

@@

We make use of the following lemma due to \citet{pilipczuk.siebertz:polynomial-arxiv,pilipczuk.siebertz:polynomial-soda}
\begin{lem}\lemlabel{p-centered-treewidth}
  Any graph $H$ of treewidth at most $t$ has $\chi_p(H)\in O(p^t)$.
\end{lem}

@@

    \item $\gamma(v)=\gamma(x)$ where $\gamma:V(H)\to\{1,\ldots,O(p^t)\}$ is a $p$-centered colouring of $H$.
  \end{enumerate}
  Condition 1 is easy to satisfy since $|V_i\cap B_x|\le \ell$.  Condition~2 is straightforward.  Condition~3 is possible to satisfy thanks to \lemref{p-centered-treewidth}.  To see why this is a $p$-centered colouring, consider some connected subgraph $X\subseteq G$.
  \begin{enumerate}
    \item If there exists $v,w\in V(X)$ with $v\in V_i$ and $w\in V_j$ with $j-i\ge p$ then, since $G[X]$ is connected, $G[X]$ contains a path from $v$ to $w$.  By the definition of layering, this path contains at least one vertex from $V_{i'}$ for each $i'\in\{i,i+1,\ldots,j\}$. Therefore, $|\{\beta(v'):v'\in X\}|\ge j-i+1 > p$, so $X$ receives more than $p$ distinct colours.

    \item Otherwise, $X\subseteq V_{i},\ldots,V_{i+s}$ for some $s<p$.  Let $H'=H[\{x\in V(H):B_x\cap X\neq\emptyset]$.  If $|\{\gamma(x):x\in V(H')\}| > p$ then $|\{\gamma(v):v\in X\}|> p$ so $|\{\phi(v):v\in X\}|> p$ and we are done.  Otherwise, since $\gamma$ is a $p$-centered colouring of $H$, there must exist some $x\in V(H')$ such that $\gamma(x)\neq\gamma(y)$ for every $y\in V(H')\setminus\{x\}$.
    For any $v,w\in B_x$ with $v\neq w$, either $v,w\in V_{i'}$ for some $i'\in\{i,i+1,\ldots,i+s\}$ in which case $\alpha(v)\neq\alpha(w)$; or $v\in V_{i'}$ and $w\in V_{i''}$ with $|i'-i''|< p$, in which case $\beta(v)\neq\beta(w)$. Therefore every vertex $v\in B_x$ receives a colour $\phi(v)$ distinct from every colour in $\{\phi(z):z\in X\setminus\{x\}\}$. Therefore, every vertex in $B_x$ receives colour distinct from every other vertex in $X$.
  \end{enumerate}
\end{proof}

@@

Prior to the current work, the strongest upper bounds on the number of colours required for $p$-centered colourings of $k$-planar graphs used the fact that $k$-planar graphs have bounded expansion.  \note{PM}{I didn't follow the discussion on the previous best-known bounds for $p$-centered colourings.  Can someone fill me in?}

\subsection{More Stuff}

\begin{enumerate}
  \item clustered colouring;
  \item connection with queue layouts;
  \item $d$-map graphs, which are $d^2$-planar;
  \item connection with low-treewidth/treedepth colourings;
  \item An important open question is whether $k$-planar graphs have layered $H$-partitions whose layered width is upper bounded by a function of $k$ but in which the treewidth of $H$ is bounded by an absolute constant, independent of $k$.
  \item Are our current $H$-partitions $K_{3,s}$-free for some $s=s(k)$?  If so, this is good news for clustered colouring.
\end{enumerate}
